---
title: |
  <center> Statistical Research Skills </center>
  <center> Group 6 </center>
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document:
    df_print: paged
  pdf_document: default
---
## (1) Load required packages  

```{r}
library(terra)
library(data.table)
library(spdep)
library(INLA)
library(scoringRules)
library(ncdf4)
library(geodata)
library(ggplot2)
```
---


# (1) create ds_us (1901-2024).nc


# (1) Create df_usa.nc and elev_usa.tif

```{r}

r_temp  <- rast("ds_us (1901-2024).nc", subds = "tmp")

# add elevation data
elev <- elevation_30s(country = "USA", path = "data/")

elev_05deg <- resample(elev, r_temp[[1]], method = "average")

# combine layers of data
r_all <- c(r_temp)

writeCDF(r_all, "ds_usa.nc", overwrite=TRUE)
writeRaster(elev_05deg, "elev_usa.tif", overwrite = TRUE)
```


# (2) Read Input Data for Pre & Processing

```{r}

nc_file   <- "ds_usa.nc"
elev_file <- "elev_usa.tif"

r_temp <- rast(nc_file, subds = "tmp")

stopifnot(inherits(r_temp, "SpatRaster"))
stopifnot(nlyr(r_temp) >= 2) 

ref <- r_temp[[1]]
if (is.na(crs(ref))) stop("temp raster has no CRS; cannot align covariates safely.")

r_elev <- rast(elev_file)

if (!same.crs(r_elev, ref)) r_elev <- project(r_elev, ref)

r_elev <- resample(r_elev, ref, method = "bilinear")

r_elev <- r_elev[[1]]

```

---

# (3) Create a time index and label layers

```{r}
full_start <- as.Date("1901-01-01")
tt_full <- seq(full_start, by="month", length.out = nlyr(r_temp))
keep_layers <- which(tt_full >= as.Date("1901-01-01"))
r_temp <- r_temp[[keep_layers]]
tt <- tt_full[keep_layers]
names(r_temp) <- format(tt, "%Y-%m")
```

---

# (4) Reshape data from raster to long format (cell × year)

```{r}
# convert r_temp to season temperature ranges

df_temp <- as.data.frame(r_temp, xy = TRUE, cells = TRUE, na.rm = FALSE)


dt <- melt(as.data.table(df_temp),
           id.vars = c("cell","x","y"),
           variable.name = "ym",
           value.name = "tmp")

dt[, year  := as.integer(substr(ym, 1, 4))]
dt[, month := as.integer(substr(ym, 6, 7))]

dt <- dt[!is.na(tmp)]

dt_range <- dt[, .(tmp_range = max(tmp, na.rm = TRUE) - min(tmp, na.rm = TRUE)),
               by = .(x, y, cell, year)]

setnames(dt_range, c("x","y"), c("lon","lat"))

```
---

# (5) Merge static covariates (robust join by coordinates)

```{r}

df_elev <- as.data.frame(r_elev, xy = TRUE, na.rm = FALSE)
df_elev <- as.data.table(df_elev)

setnames(df_elev, names(df_elev)[3], "elev")

df_elev[, `:=`(x = round(x, 6), y = round(y, 6))]
dt_range[, `:=`(lon = round(lon, 6), lat = round(lat, 6))]

dt_range <- merge(
  dt_range,
  df_elev,
  by.x = c("lon","lat"),
  by.y = c("x","y"),
  all.x = TRUE
)

dt_range <- dt_range[!is.na(elev)]

dt_range[, elev_sc := as.numeric(scale(elev))]

stopifnot(nrow(dt_range) > 0)

dt_range[, area_id := as.integer(factor(cell))]
dt_range[, cell := NULL]

dt_range$t_id <- as.integer(factor(dt_range$year))

```

---

# (6) Create spatial index

```{r}

dt_inla <- copy(dt_range)

S <- max(dt_inla$area_id)
```

---

# (7) Construct spatial neighbourhood graph (using rook adjacency)


```{r}
cells <- unique(dt_inla[, .(area_id, lon, lat)])
setorder(cells, lat, lon)

ux <- sort(unique(cells$lon))
uy <- sort(unique(cells$lat))
ncol <- length(ux)
nrow <- length(uy)

cells[, col := match(lon, ux)]
cells[, row := match(lat, uy)]

mat <- matrix(NA_integer_, nrow = nrow, ncol = ncol)
mat[cbind(cells$row, cells$col)] <- cells$area_id
idx_to_area <- as.vector(mat)

nb_full <- cell2nb(nrow, ncol, type = "rook")
nb_area <- vector("list", length = S)

for (idx in seq_along(nb_full)) {
  a <- idx_to_area[idx]
  if (is.na(a)) next
  neigh_idx  <- nb_full[[idx]]
  neigh_area <- idx_to_area[neigh_idx]
  neigh_area <- neigh_area[!is.na(neigh_area)]
  nb_area[[a]] <- sort(unique(as.integer(neigh_area)))
}

for (i in seq_len(S)) if (is.null(nb_area[[i]])) nb_area[[i]] <- integer(0)

write_inla_graph <- function(nb, file) {
  S <- length(nb)
  con <- file(file, open = "wt")
  on.exit(close(con), add = TRUE)
  writeLines(as.character(S), con)
  for (i in seq_len(S)) {
    neigh <- nb[[i]]
    if (is.null(neigh)) neigh <- integer(0)
    neigh <- sort(unique(as.integer(neigh)))
    writeLines(paste(c(i, length(neigh), neigh), collapse = " "), con)
  }
}

graph_file <- file.path(tempdir(), "usa_grid.graph")
write_inla_graph(nb_area, graph_file)
g <- inla.read.graph(graph_file)

```

---


# (8) Fit candidate INLA models (Gaussian likelihood)

```{r}

ctrl_pred <- list(compute = TRUE, link = 1)

ctrl_comp <- list(
  dic = TRUE,
  waic = TRUE,
  cpo = FALSE,
  config=TRUE
)


inla.setOption(num.threads = 8)

# M0: fixed effects only (no spatial/temporal structure)
m0 <- inla(
  tmp_range ~ 1 + elev_sc,
  data = dt_inla,
  family = "lognormal",
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M1: + temporal RW2 trend
m1 <- inla(
  tmp_range ~ 1 + elev_sc +
    f(t_id, model = "rw2", constr = TRUE),
  data = dt_inla,
  family = "lognormal",
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)


# M2: + spatial BYM2 effect
m2 <- inla(
  tmp_range ~ 1 + elev_sc +
    f(area_id, model = "bym2", graph = g,
      scale.model = TRUE, constr = TRUE),
  data = dt_inla,
  family = "lognormal",
  # control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M3: full spatio-temporal model
m3 <- inla(
  tmp_range ~ 1 + elev_sc +
    f(t_id, model = "rw2", constr = TRUE) +
    f(area_id, model = "bym2", graph = g,
      scale.model = TRUE, constr = TRUE),
  data = dt_inla,
  family = "lognormal",
  # control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

```


---


# (9) Model comparison


```{r}

cat("M0 WAIC =",  m0$waic$waic, "\n")
cat("M1 WAIC =",  m1$waic$waic, "\n")
cat("M2 WAIC =",  m2$waic$waic, "\n")
cat("M3 WAIC =",  m3$waic$waic)

```

# (10) Model summaries


```{r}


m3$summary.fixed


m3$summary.hyperpar


```




# (11) Absolute Goodness of Fit Checks (posterior predictive checks, etc.)


```{r}
# Posterior predictive checks:

# sample for model
n_sims <- 100
y_rep <- matrix(NA, nrow = nrow(dt_inla), ncol = n_sims)
post_samples <- inla.posterior.sample(n_sims, m3)
for(i in 1:n_sims){
  s <- post_samples[[i]]
  eta <- s$latent[1:nrow(dt_inla)]                  
  sigma <- 1 / sqrt(s$hyperpar[[1]])
  y_rep[,i] <- rlnorm(nrow(dt_inla), meanlog = eta, sdlog = sigma)
}
y_rep_all <- as.vector(y_rep)

```



```{r}
# Residual diagnostics using bayesian studentized residuals:

# compute residuals
n <- nrow(dt_range)
y <- log(dt_range$tmp_range)
X <- cbind(1, dt_range$elev_sc)
XtX_inv <- solve(t(X) %*% X)
h <- rowSums((X %*% XtX_inv) * X)

studentisedred <- matrix(NA, nrow = n, ncol = n_sims)
for(l in 1:n_sims){
  s <- post_samples[[l]]
  eta <- s$latent[1:n]                   
  sigma <- 1 / sqrt(s$hyperpar[[1]])
  studentisedred[,l] <- (y - eta) / sqrt(sigma^2 * (1 - h))
}
studentisedredm <- apply(studentisedred, 1, mean)


# normality (qq plot) - checks log normality of data
qqnorm(studentisedredm,
       main="",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       pch = 16, cex = 0.4, col = "black",
       xlim = c(-3,3), ylim = c(-3,3),
       cex.lab = 1.25)
qqline(studentisedredm, col = "red", lwd = 1.5)

```



```{r}

# more numeric absolute measures of fit

# bayesian posterior predictive p-values of mean and variance

yrepmean <- apply(y_rep, 2, mean)
yrepvar  <- apply(y_rep, 2, var)

obs_mean <- mean(dt_inla$tmp_range)
obs_var <- var(dt_inla$tmp_range)

p_mean <- mean(yrepmean >= obs_mean)
p_var  <- mean(yrepvar >= obs_var)

bayes_pvals <- c(
                 mean = p_mean,
                 variance = p_var)

print(bayes_pvals)

# crps
crps_values <- crps_sample(y = dt_range$tmp_range, dat = y_rep)
crps_mean <- mean(crps_values)

cat("CRPS =", round(crps_mean, 3), "\n")


```




# (12) Map


```{r}

spatial <- m3$summary.random$area_id
spatial_df <- data.frame(spatial$ID, spatial$mean, spatial$sd)
colnames(spatial_df) <- c("area_id", "spatial_mean", "spatial_sd")

df_spatial <- merge(dt_inla, spatial_df, by="area_id", all.x=TRUE)

cor(df_spatial$spatial_mean, df_spatial$elev)

spatial_cells <- df_spatial[!duplicated(df_spatial$area_id), ]

nrow(spatial_cells)


```

```{r}


r <- rast(
  spatial_cells[, c("lon", "lat", "spatial_mean")],
  type = "xyz",
  crs = "EPSG:4326"
)

r_df <- as.data.frame(r, xy = TRUE)

ggplot(spatial_cells) +
  geom_raster(aes(x = lon, y = lat, fill = spatial_mean)) +
  coord_fixed() +
  scale_fill_gradient2(low="#2B4220", mid="#40821f", high="khaki1", midpoint=-0.5) +
  theme_classic() +
  theme(panel.border=element_rect(colour="black", fill=NA)) +
  labs(
    title = "",
    fill = "Effect"
  ) +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size=14)
  ) +
  scale_x_continuous(expand = c(0.01,0.01)) +
  scale_y_continuous(expand = c(0.01,0.01))


ggsave("spatial.png", width=8, height=5, dpi=300)


```






