---
title: |
  <center> Statistical Research Skill </center>
  <center> Group 6 </center>
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Model Structure

The Bayesian hierarchical spatio-temporal (BHST) model is specified as:

$$
y_{s,t}
=
\mu
+
\beta_{1}\,\text{Elev}_s
+
\beta_{2}\,\text{DistCoast}_s
+
u_s
+
\delta(t)
+
\epsilon_{s,t}
$$

where:

- $y_{s,t}$ denotes the observed temperature anomaly at spatial location $s$ and time $t$.

**Model Components**

- **Global mean ($\mu$)**  
  Represents the overall average temperature level across all time periods and spatial locations.

- **Elevation effect ($\beta_1 \text{Elev}_s$)**  
  Captures the influence of elevation on temperature variability.  
  Higher elevation areas are typically associated with lower temperatures.

- **Distance-to-coast effect ($\beta_2 \text{DistCoast}_s$)**  
  Represents the influence of proximity to the coastline.  
  Coastal regions often experience moderated temperatures compared to inland areas.

- **Smooth temporal trend ($\delta(t)$)**  
  Captures long-term climate evolution over time.  
  A second-order random walk (RW2) prior is used to allow a flexible, smooth nonlinear trend without imposing a strict linear structure.

- **Spatial structured effect ($u_s$)**  
  Models spatial dependence between neighbouring grid cells.  
  This component is specified using a Conditional Autoregressive (CAR) or BYM2 prior, allowing geographically nearby locations to exhibit similar temperature patterns.

- **Observation error ($\epsilon_{s,t}$)**  
  Represents unexplained variability, assumed to follow a Gaussian distribution:

$$
\epsilon_{s,t} \sim \mathcal{N}(0,\sigma_\epsilon^2)
$$

$$
\epsilon_{s,t} \sim \mathcal{N}(0,\sigma_{\epsilon}^{2})
$$

The model decomposes temperature variability into covariate effects (elevation and distance to coast), a smooth long-term temporal trend, spatially structured variability across grid cells, and residual noise.

# RCode

## (1) Load required packages  
Here, we used package `terra` for raster/NetCDF handling, `data.table` for fast reshaping and joins, `spdep` for neighbourhood structure, and `INLA` for Bayesian inference. 
```{r}
library(terra)
library(data.table)
library(spdep)
library(INLA)
```


## (2) Read Input Data for Pre & Processing
Dataset used 
(1) Monthly anomaly raster `anom` from `ds_usa.nc`.  
(2) Static covariates from GeoTIFF:
     - elevation (`elev_usa.tif`)
     - distance to coast (`d2c_usa.tif`)
(3) Preprocessing : Align covariates to the anomaly grid using CRS transformation and `resample() i.e. make standard spatial grid for both (covariates and anomalies)
```{r}
# (1) Input File ----------------------------------------------------------
nc_file   <- "ds_usa.nc"
elev_file <- "elev_usa.tif"
d2c_file  <- "d2c_usa.tif"

# (2) Input and Preprocessing Raster Data ---------------------------------
# TEMPERATURE DATASETS ----------------------------------------------------
# Load temperature anomaly data from NetCDF
# 'anom' contains monthly anomaly layers (time-varying)
r_anom <- rast(nc_file, subds = "anom")

# checking format r_anom
stopifnot(inherits(r_anom, "SpatRaster")) #ensure r_anom is a SpatRaster
stopifnot(nlyr(r_anom) >= 2) #ensure r_anom contains multiple layer

# Get the first anomaly layer as the reference grid
ref <- r_anom[[1]]
# Check that ref has a coordinate reference system
if (is.na(crs(ref))) stop("anom raster has no CRS; cannot align covariates safely.")

# COVARIATE DATASETS ----------------------------------------------------
# Now, load the static covariates, elevation and distance to coast
r_elev <- rast(elev_file) #elevation
r_d2c  <- rast(d2c_file) #distance to coast

# Adjust spatial grid covariates to temperature anomalies (same CRS & res)
# Ensure elevation has the same coordinate reference system as the anomaly grid (use ref)
# Reproject to match the reference raster if it's different
if (!same.crs(r_elev, ref)) r_elev <- project(r_elev, ref)

# Resample elevation so that its resolution and grid alignment
r_elev <- resample(r_elev, ref, method = "bilinear")

# get only single elevation layer
r_elev <- r_elev[[1]]

# Same procedure, but in distance to coast
if (!same.crs(r_d2c, ref)) r_d2c <- project(r_d2c, ref)
r_d2c <- resample(r_d2c, ref, method = "bilinear")
r_d2c <- r_d2c[[1]]

```

# (3) Create a time index and label layers  
- Build a monthly date sequence starting from `1961-01-01`
- Assign layer names in `YYYY-MM` format for consistent reshaping  
- Subset layers by year range using `keep_layers`
```{r}
full_start <- as.Date("1961-01-01")
tt_full <- seq(full_start, by="month", length.out = nlyr(r_anom))
keep_layers <- which(tt_full >= as.Date("1961-01-01")) #here to subset the years
r_anom <- r_anom[[keep_layers]]
tt <- tt_full[keep_layers]
names(r_anom) <- format(tt, "%Y-%m")
```


# (4) Reshape data from raster to long format (cell × month) 
- Convert raster to a wide data frame with `as.data.frame(..., cells=TRUE, xy=TRUE)`.  
- Convert to long format using `melt()` into columns:
- spatial identifiers (`cell`, `lon`, `lat`)
- time identifier (`ym`, `t_id`)
- response (`y_obs`)
```{r}
df_anom <- as.data.frame(r_anom, xy = TRUE, cells = TRUE, na.rm = FALSE)

dt <- melt(as.data.table(df_anom),
           id.vars = c("cell","x","y"),
           variable.name = "ym", value.name = "anom")

setnames(dt, old = c("x","y"), new = c("lon","lat"))

dt[, year  := as.integer(substr(ym, 1, 4))]
dt[, month := as.integer(substr(ym, 6, 7))]

ym_levels <- sort(unique(dt$ym))
dt[, t_id := match(ym, ym_levels)]

# response
dt[, y_obs := anom]
```

## (5) Filter valid spatial domain  
Drop grid cells that are missing (`NA`) for **all months** to avoid including ocean/out-of-domain cells
```{r}
cell_ok <- dt[, .(keep = any(!is.na(y_obs))), by = cell][keep == TRUE, cell]
dt <- dt[cell %in% cell_ok]
```

## (6) Merge static covariates (robust join by coordinates) 
- Convert elevation and distance-to-coast rasters to `(x,y,value)` tables.
- Round coordinates to avoid floating-point mismatch.Coordinates were rounded to six decimal places in order to avoid floating-point precision issues during spatial joins. References : Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. R Journal 
- Merge covariates into the long table using `(lon, lat)` join (not `cell`).  
- Scale covariates:
  - `elev_sc = scale(elev)`
  - `d2c_sc = scale(log1p(dist_to_coast))`

```{r}
df_elev <- as.data.frame(r_elev, xy = TRUE, na.rm = FALSE)  # x y value
df_d2c  <- as.data.frame(r_d2c,  xy = TRUE, na.rm = FALSE)

# convert to data.table BEFORE using :=
df_elev <- as.data.table(df_elev)
df_d2c  <- as.data.table(df_d2c)

# rename value columns safely (3rd column is value)
setnames(df_elev, names(df_elev)[3], "elev")
setnames(df_d2c,  names(df_d2c)[3],  "dist_to_coast")

# round coords to avoid floating-point join issues
df_elev[, `:=`(x = round(x, 6), y = round(y, 6))]
df_d2c[,  `:=`(x = round(x, 6), y = round(y, 6))]
dt[,      `:=`(lon = round(lon, 6), lat = round(lat, 6))]

# merge covariates then join to dt
covs_xy <- merge(df_elev, df_d2c, by = c("x", "y"), all = TRUE)
dt <- merge(dt, covs_xy, by.x = c("lon","lat"), by.y = c("x","y"), all.x = TRUE)

# drop rows missing covariates (avoid silent drops inside INLA)
dt <- dt[!is.na(elev) & !is.na(dist_to_coast)]

# scale covariates
dt[, elev_sc := as.numeric(scale(elev))]
dt[, d2c_sc  := as.numeric(scale(log1p(dist_to_coast)))]

# reindex spatial after filtering
stopifnot(nrow(dt) > 0)
dt[, area_id := as.integer(factor(cell))]
```


## (7) Create spatial index and train/test split 
- Define spatial ID: `area_id = factor(cell)` after filtering
- Split the datasets:
  - Train: `year <= 2020`
  - Test: `year > 2020`
- Create `dt_inla` for prediction by setting `y_obs = NA` in the test set

```{r}
dt[, set := fifelse(year <= 2020, "train", "test")]

dt_truth <- copy(dt)

dt_inla <- copy(dt)
dt_inla[set == "test", y_obs := NA_real_]

S <- max(dt_inla$area_id)
```

## (8) Construct spatial neighbourhood graph (using rook adjacency)  
- Arrange cells on the grid using unique `lon/lat`  
- Build rook adjacency using `cell2nb()` 
- Convert to an INLA graph file and read it with `inla.read.graph()`
```{r}
cells <- unique(dt_inla[, .(area_id, lon, lat)])
setorder(cells, lat, lon)

ux <- sort(unique(cells$lon))
uy <- sort(unique(cells$lat))
ncol <- length(ux)
nrow <- length(uy)

cells[, col := match(lon, ux)]
cells[, row := match(lat, uy)]

mat <- matrix(NA_integer_, nrow = nrow, ncol = ncol)
mat[cbind(cells$row, cells$col)] <- cells$area_id
idx_to_area <- as.vector(mat)

nb_full <- cell2nb(nrow, ncol, type = "rook")
nb_area <- vector("list", length = S)

for (idx in seq_along(nb_full)) {
  a <- idx_to_area[idx]
  if (is.na(a)) next
  neigh_idx  <- nb_full[[idx]]
  neigh_area <- idx_to_area[neigh_idx]
  neigh_area <- neigh_area[!is.na(neigh_area)]
  nb_area[[a]] <- sort(unique(as.integer(neigh_area)))
}
for (i in seq_len(S)) if (is.null(nb_area[[i]])) nb_area[[i]] <- integer(0)

write_inla_graph <- function(nb, file) {
  S <- length(nb)
  con <- file(file, open = "wt")
  on.exit(close(con), add = TRUE)
  writeLines(as.character(S), con)
  for (i in seq_len(S)) {
    neigh <- nb[[i]]
    if (is.null(neigh)) neigh <- integer(0)
    neigh <- sort(unique(as.integer(neigh)))
    writeLines(paste(c(i, length(neigh), neigh), collapse = " "), con)
  }
}

graph_file <- file.path(tempdir(), "usa_grid.graph")
write_inla_graph(nb_area, graph_file)
g <- inla.read.graph(graph_file)
```

## (9) Fit candidate INLA models (Gaussian likelihood)  
- **M0:** intercept-only baseline  
- **M1:** baseline + RW2 temporal trend  
- **M2:** M1 + BYM2 spatial effect (graph-based)  
- **M3:** M2 + covariates (`elev_sc`, `d2c_sc`)  
All models are fitted as Gaussian:
$$
y_{s,t} \sim \mathcal{N}(\eta_{s,t}, \sigma_\epsilon^2).
$$

```{r}
ctrl_pred <- list(compute = TRUE, link = 1)
ctrl_comp <- list(dic = TRUE, waic = TRUE, cpo = FALSE)
ctrl_fxd  <- list(
  mean = list("(Intercept)" = 0),
  prec = list("(Intercept)" = 1/100)   # var=100
)

inla.setOption(num.threads = 8) # setting thread to laptop cpu for faster work

# M0: fixed effects only
m0 <- inla(
  y_obs ~ 1 + elev_sc + d2c_sc,
  data = dt_inla, family = "gaussian",
  control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M1: + RW2 trend
m1 <- inla(
  y_obs ~ 1 + elev_sc + d2c_sc +
    f(t_id, model = "rw2", constr = TRUE),
  data = dt_inla, family = "gaussian",
  control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M2: + BYM2 spatial
m2 <- inla(
  y_obs ~ 1 + elev_sc + d2c_sc +
    f(area_id, model = "bym2", graph = g, scale.model = TRUE, constr = TRUE),
  data = dt_inla, family = "gaussian",
  control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M3: + RW2 trend + BYM2 spatial
m3 <- inla(
  y_obs ~ 1 + elev_sc + d2c_sc +
    f(t_id, model = "rw2", constr = TRUE) +
    f(area_id, model = "bym2", graph = g, scale.model = TRUE, constr = TRUE),
  data = dt_inla, family = "gaussian",
  control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)
```

## (10) Model comparison and predictive evaluation (2021–2024)**  
- Compare models using:
  - **WAIC**
  - **DIC**
- Compute test performance on the hold-out period using:
  - **RMSE**
  - **MAE**
- Report fixed effects from the final covariate model via `summary.fixed`
```{r}
get_metrics <- function(mod){
  data.table(DIC = mod$dic$dic, WAIC = mod$waic$waic)
}

score_test <- function(mod, dt_inla, dt_truth){
  idx_test <- which(dt_inla$set == "test")
  pred  <- mod$summary.linear.predictor$mean[idx_test]
  truth <- dt_truth$y_obs[idx_test]
  ok <- !is.na(truth) & !is.na(pred)
  pred <- pred[ok]; truth <- truth[ok]
  data.table(
    n_test = length(truth),
    RMSE = sqrt(mean((pred - truth)^2)),
    MAE  = mean(abs(pred - truth))
  )
}

mods <- list(M0 = m0, M1 = m1, M2 = m2, M3 = m3)

metrics <- rbindlist(lapply(names(mods), function(nm){
  cbind(model = nm, get_metrics(mods[[nm]]))
}))
test_scores <- rbindlist(lapply(names(mods), function(nm){
  cbind(model = nm, score_test(mods[[nm]], dt_inla, dt_truth))
}))

summary(m0)
summary(m1)
summary(m2)
summary(m3)

print(metrics[order(WAIC)])
print(test_scores[order(RMSE)])
```

