---
title: "data_setup"
output: html_document
date: "2026-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up predictor variable (annual temperature range)

now we're gonna set up our data using max temperature range


```{r}
library(terra)

temp <- rast("us_temp.nc", subds="tmp")

time_vals <- time(temp)

years <- format(time_vals, "%Y")

temp_range <- tapp(temp, years, function(x) {
  max(x) - min(x)
})


```

## Read in distance to coastline data

```{r}
r_d2c <- rast("us_temp.nc", subds="distance_to_coast")
```


## Calculate elevation for our grids

```{r}

install.packages("geodata")
install.packages("ncdf4")

library(ncdf4)
library(terra)
library(geodata)

# add elevation data
elev <- elevation_30s(country = "USA", path = "data/")
plot(elev)

elev_05deg <- resample(elev, temp[[1]], method = "average")
plot(elev_05deg)
```


## Combine all elements into one raster file

```{r}

# combine layers of data
r_all <- c(temp_range, elev_05deg, r_d2c)
year_names <- paste0("temp.range.", 1901:2012)
names(r_all)[1:nlyr(temp_range)] <- year_names 
names(r_all)[(nlyr(temp_range)+1)] <- "elev"
names(r_all)[(nlyr(temp_range)+2)] <- "dist_to_coast"

writeCDF(r_all, "ds_usa.nc", overwrite=TRUE)

```


## Check raster file has been exported successfully
our data goes from Jan 1901 to December 2012, so we should have 112 layers for 
temperature range, one layer for elevation and one for distance to coastline

```{r}

data_rast <- rast("ds_usa.nc")

data_rast

```

```{r}

df <- as.data.frame(data_rast, xy = TRUE, na.rm = TRUE)

df
```



