---
title: |
  <center> Statistical Research Skill </center>
  <center> Group 6 </center>
subtitle: School of Mathematics, University of Edinburgh
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Model Structure

The Bayesian hierarchical spatio-temporal (BHST) model is specified as:

$$
y_{s,t}
=
\mu
+
\beta_{1}\,\text{Elev}_s
+
\beta_{2}\,\text{DistCoast}_s
+
u_s
+
\delta(t)
+
\epsilon_{s,t}
$$

where:

- $y_{s,t}$ denotes the **annual temperature range** at spatial location $s$ and time $t$.

**Model Components**

- **Global mean ($\mu$)**  
  Represents the overall average temperature range across all time periods and spatial locations.

- **Elevation effect ($\beta_1 \text{Elev}_s$)**  
  Captures the influence of elevation on temperature variability.  
  Higher elevation areas often experience stronger temperature variation.

- **Distance-to-coast effect ($\beta_2 \text{DistCoast}_s$)**  
  Represents the influence of proximity to the coastline.  
  Coastal regions often experience moderated temperatures compared to inland areas.

- **Smooth temporal trend ($\delta(t)$)**  
  Captures long-term climate evolution over time.  
  A second-order random walk (RW2) prior is used to allow a flexible, smooth nonlinear trend without imposing a strict linear structure.

- **Spatial structured effect ($u_s$)**  
  Models spatial dependence between neighbouring grid cells.  
  This component is specified using a Conditional Autoregressive (CAR) or BYM2 prior, allowing geographically nearby locations to exhibit similar temperature patterns.

- **Observation error ($\epsilon_{s,t}$)**  
  Represents unexplained variability, assumed to follow a Gaussian distribution:

$$
\epsilon_{s,t} \sim \mathcal{N}(0,\sigma_\epsilon^2)
$$

The model decomposes temperature range variability into covariate effects (elevation and distance to coast), a smooth long-term temporal trend, spatially structured variability across grid cells, and residual noise.

---

# RCode

## (1) Load required packages  
Here, we used package `terra` for raster/NetCDF handling, `data.table` for fast reshaping and joins, `spdep` for neighbourhood structure, and `INLA` for Bayesian inference. 

```{r}
library(terra)
library(data.table)
library(spdep)
library(INLA)
```
---

# (2) Read Input Data for Pre & Processing

Dataset used  

(1) **Annual temperature range raster** from `ds_usa.nc`  
(2) Static covariates from GeoTIFF:

- elevation (`elev_usa.tif`)
- distance to coast (`d2c_usa.tif`)

(3) Preprocessing : Align covariates to the temperature grid using CRS transformation and `resample()` i.e. make standard spatial grid for both (covariates and temperature range)

```{r}
# (1) Input File ----------------------------------------------------------

nc_file   <- "ds_usa.nc"
elev_file <- "elev_usa.tif"
d2c_file  <- "d2c_usa.tif"

# (2) Input and Preprocessing Raster Data ---------------------------------
# TEMPERATURE DATASETS ----------------------------------------------------
# Load temperature data from NetCDF
r_temp <- rast(nc_file, subds = "tmp")

# checking format r_temp
stopifnot(inherits(r_temp, "SpatRaster")) #ensure r_temp is a SpatRaster
stopifnot(nlyr(r_temp) >= 2) #ensure r_temp contains multiple layer

# Get the first temp layer as the reference grid
ref <- r_temp[[1]]
# Check that ref has a coordinate reference system
if (is.na(crs(ref))) stop("temp raster has no CRS; cannot align covariates safely.")

# COVARIATE DATASETS ----------------------------------------------------
# Now, load the static covariates, elevation and distance to coast
r_elev <- rast(elev_file) #elevation
r_d2c  <- rast(d2c_file) #distance to coast

# Adjust spatial grid covariates to temperature (same CRS & res)
# Ensure elevation has the same coordinate reference system as the temp grid (use ref)
# Reproject to match the reference raster if it's different
if (!same.crs(r_elev, ref)) r_elev <- project(r_elev, ref)

# Resample elevation so that its resolution and grid alignment
r_elev <- resample(r_elev, ref, method = "bilinear")

# get only single elevation layer
r_elev <- r_elev[[1]]

# Same procedure, but in distance to coast
if (!same.crs(r_d2c, ref)) r_d2c <- project(r_d2c, ref)
r_d2c <- resample(r_d2c, ref, method = "bilinear")
r_d2c <- r_d2c[[1]]
```

---

# (3) Create a time index and label layers

- Extract year labels from raster layer names
- Assign layer names in `YYYY` format for consistent reshaping

```{r}
full_start <- as.Date("1901-01-01")
tt_full <- seq(full_start, by="month", length.out = nlyr(r_temp))
keep_layers <- which(tt_full >= as.Date("1901-01-01")) #here to subset the years
r_temp <- r_temp[[keep_layers]]
tt <- tt_full[keep_layers]
names(r_temp) <- format(tt, "%Y-%m")
```

---

# (4) Reshape data from raster to long format (cell × year)

- Convert raster to a wide data frame with `as.data.frame(..., cells=TRUE, xy=TRUE)`
- Convert to long format using `melt()`
- note that by doing this, we will have many NA temperatures. these obs. are not
in mainland US and will be dropped in the next step

```{r}
# convert r_temp to season temperature ranges

df_temp <- as.data.frame(r_temp, xy = TRUE, cells = TRUE, na.rm = FALSE)


dt <- melt(as.data.table(df_temp),
           id.vars = c("cell","x","y"),
           variable.name = "ym",
           value.name = "tmp")

dt[, year  := as.integer(substr(ym, 1, 4))]
dt[, month := as.integer(substr(ym, 6, 7))]

dt <- dt[!is.na(tmp)]

dt_range <- dt[, .(tmp_range = max(tmp, na.rm = TRUE) - min(tmp, na.rm = TRUE)),
               by = .(x, y, cell, year)]

setnames(dt_range, c("x","y"), c("lon","lat"))

```

---

# (5) Filter valid spatial domain  

Drop grid cells that are missing (`NA`) for **all years** to avoid including ocean/out-of-domain cells.

```{r}
# cell_ok <- dt[, .(keep = any(!is.na(y_obs))), by = cell][keep == TRUE, cell]
# dt <- dt[cell %in% cell_ok]
```

---

# (6) Merge static covariates (robust join by coordinates)

```{r}
df_elev <- as.data.frame(r_elev, xy = TRUE, na.rm = FALSE)
df_d2c  <- as.data.frame(r_d2c,  xy = TRUE, na.rm = FALSE)

df_elev <- as.data.table(df_elev)
df_d2c  <- as.data.table(df_d2c)

setnames(df_elev, names(df_elev)[3], "elev")
setnames(df_d2c,  names(df_d2c)[3],  "dist_to_coast")

df_elev[, `:=`(x = round(x, 6), y = round(y, 6))]
df_d2c[,  `:=`(x = round(x, 6), y = round(y, 6))]
dt_range[,      `:=`(lon = round(lon, 6), lat = round(lat, 6))]

covs_xy <- merge(df_elev, df_d2c, by = c("x", "y"), all = TRUE)

dt_range <- merge(dt_range, covs_xy,
            by.x = c("lon","lat"),
            by.y = c("x","y"),
            all.x = TRUE)

dt_range <- dt_range[!is.na(elev) & !is.na(dist_to_coast)]

dt_range[, elev_sc := as.numeric(scale(elev))]
dt_range[, d2c_sc  := as.numeric(scale(log1p(dist_to_coast)))]

stopifnot(nrow(dt_range) > 0)

dt_range[, area_id := as.integer(factor(cell))]


# get rid of unneeded/duplicate columns
dt_range <- dt_range[,-c("cell")]

# add temporal index

dt_range$t_id <- as.integer(factor(dt_range$year)) # make temporal index


```

---

# (7) Create spatial index

```{r}
# (7) Create spatial index

dt_inla <- copy(dt_range)

# number of spatial areas
S <- max(dt_inla$area_id)
```

---

# (8) Construct spatial neighbourhood graph (using rook adjacency)


```{r}
cells <- unique(dt_inla[, .(area_id, lon, lat)])
setorder(cells, lat, lon)

ux <- sort(unique(cells$lon))
uy <- sort(unique(cells$lat))
ncol <- length(ux)
nrow <- length(uy)

cells[, col := match(lon, ux)]
cells[, row := match(lat, uy)]

mat <- matrix(NA_integer_, nrow = nrow, ncol = ncol)
mat[cbind(cells$row, cells$col)] <- cells$area_id
idx_to_area <- as.vector(mat)

nb_full <- cell2nb(nrow, ncol, type = "rook")
nb_area <- vector("list", length = S)

for (idx in seq_along(nb_full)) {
  a <- idx_to_area[idx]
  if (is.na(a)) next
  neigh_idx  <- nb_full[[idx]]
  neigh_area <- idx_to_area[neigh_idx]
  neigh_area <- neigh_area[!is.na(neigh_area)]
  nb_area[[a]] <- sort(unique(as.integer(neigh_area)))
}

for (i in seq_len(S)) if (is.null(nb_area[[i]])) nb_area[[i]] <- integer(0)

write_inla_graph <- function(nb, file) {
  S <- length(nb)
  con <- file(file, open = "wt")
  on.exit(close(con), add = TRUE)
  writeLines(as.character(S), con)
  for (i in seq_len(S)) {
    neigh <- nb[[i]]
    if (is.null(neigh)) neigh <- integer(0)
    neigh <- sort(unique(as.integer(neigh)))
    writeLines(paste(c(i, length(neigh), neigh), collapse = " "), con)
  }
}

graph_file <- file.path(tempdir(), "usa_grid.graph")
write_inla_graph(nb_area, graph_file)
g <- inla.read.graph(graph_file)

```

---


# (9) Fit candidate INLA models (Gaussian likelihood)

```{r}

ctrl_pred <- list(compute = TRUE, link = 1)

ctrl_comp <- list(
  dic = TRUE,
  waic = TRUE,
  cpo = FALSE
)

ctrl_fxd  <- list(
  mean = list("(Intercept)" = 0),
  prec = list("(Intercept)" = 1/100)
)

inla.setOption(num.threads = 8)

# M0: fixed effects only (no spatial/temporal structure)
m0 <- inla(
  tmp_range ~ 1 + elev_sc,
  data = dt_inla,
  family = "lognormal",
  #control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M1: + temporal RW2 trend
m1 <- inla(
  tmp_range ~ 1 + elev_sc +
    f(t_id, model = "rw2", constr = TRUE),
  data = dt_inla,
  family = "gaussian",
  #control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)


# M2: + spatial BYM2 effect
m2 <- inla(
  tmp_range ~ 1 + elev_sc +
    f(area_id, model = "bym2", graph = g,
      scale.model = TRUE, constr = TRUE),
  data = dt_inla,
  family = "lognormal",
  # control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# M3: full spatio-temporal model
m3 <- inla(
  tmp_range ~ 1 + elev_sc +
    f(t_id, model = "rw2", constr = TRUE) +
    f(area_id, model = "bym2", graph = g,
      scale.model = TRUE, constr = TRUE),
  data = dt_inla,
  family = "lognormal",
  # control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)

# remove elevation and just use spatial
m4 <- inla(
  tmp_range ~ 1 +
    f(t_id, model = "rw2", constr = TRUE) +
    f(area_id, model = "bym2", graph = g,
      scale.model = TRUE, constr = TRUE),
  data = dt_inla,
  family = "lognormal",
  # control.fixed = ctrl_fxd,
  control.predictor = ctrl_pred,
  control.compute = ctrl_comp
)



m0$summary.fixed

m2$summary.fixed


0.02083577

0.03902563

```


---


# (10) Model comparison


```{r}
get_metrics <- function(mod){
  data.table(
    WAIC = mod$waic$waic
  )
}

mods <- list(
  M0 = m0,
  M1 = m1,
  M2 = m2,
  M3 = m3
)

metrics <- rbindlist(
  lapply(names(mods), function(nm){
    cbind(model = nm, get_metrics(mods[[nm]]))
  })
)

print(metrics[order(WAIC)])
```


# (11) Absolute Goodness of Fit Checks (posterior predictive checks, etc.)


```{r}

# Posterior predictive checks:





# Bayesian p-values:




# Residual diagnostics (and checking assumptions of log normality):

# extract log temps
log_tmp <- log(dt_range$temp_range)

# evaluate observed data and compare with normal
hist(log_tmp, breaks = 30, main = "Histogram of log(temp_range)", xlab = "log(temp_range)")
lines(density(log_tmp), col = "blue", lwd = 2)

# calculate residuals from fitted model
resid_log <- log(dt_range$temp_range) - m3$summary.linear.predictor$mean

# plot residuals vs mean
plot(m0$summary.fitted.values$mean, resid_log,
     xlab = "Fitted values",
     ylab = "Residuals (log scale)")
abline(h = 0, col = "red", lty = 2)

# plot residual qq plot
qqnorm(resid_log)
qqline(resid_log, col = "red", lwd = 2)



```



