## (1) Load required packages  
Here, we used package `terra` for raster/NetCDF handling, `data.table` for fast reshaping and joins, `spdep` for neighbourhood structure, and `INLA` for Bayesian inference. 

```{r}
library(terra)
library(data.table)
library(spdep)
library(INLA)
library(geodata)
```
---

# (2) Read Input Data for Pre & Processing

Dataset used  

(1) **Annual temperature range raster** from `ds_usa.nc`  
(2) Static covariates from GeoTIFF:

- elevation (`elev_usa.tif`)

(3) Preprocessing : Align covariates to the temperature grid using CRS transformation and `resample()` i.e. make standard spatial grid for both (covariates and temperature range)

```{r}
# (1) Input File ----------------------------------------------------------

nc_file   <- "ds_us (1901-2024).nc"

# (2) Input and Preprocessing Raster Data ---------------------------------
# TEMPERATURE DATASETS ----------------------------------------------------
# Load temperature data from NetCDF
r_temp <- rast(nc_file, subds = "tmp")

# checking format r_temp
stopifnot(inherits(r_temp, "SpatRaster")) #ensure r_temp is a SpatRaster
stopifnot(nlyr(r_temp) >= 2) #ensure r_temp contains multiple layer

# Get the first temp layer as the reference grid
ref <- r_temp[[1]]
# Check that ref has a coordinate reference system
if (is.na(crs(ref))) stop("temp raster has no CRS; cannot align covariates safely.")

# COVARIATE DATASETS ----------------------------------------------------
# Now, load the static covariate, elevation

# add elevation data
elev <- elevation_30s(country = "USA", path = "data/")
elev_05deg <- resample(elev, r_temp[[1]], method = "average")

r_elev <- rast(elev_05deg) #elevation


# Adjust spatial grid covariates to temperature (same CRS & res)
# Ensure elevation has the same coordinate reference system as the temp grid (use ref)
# Reproject to match the reference raster if it's different
if (!same.crs(r_elev, ref)) r_elev <- project(r_elev, ref)

# Resample elevation so that its resolution and grid alignment
r_elev <- resample(r_elev, ref, method = "bilinear")

# get only single elevation layer
r_elev <- r_elev[[1]]

```
